A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-11 sections/introduction.tex
Intro WIP

As functional languages evolved the need for more expressive
power has motivated language designers to look beyond Hindley-Milner.
In particular one popular direction has been to allow 
\emph{higher-ranked polymorphism} where polymorphic types can
occur anywhere in a type signature.  
An important challenge is that full type inference for higher-ranked
polymorphism is know to be undecidable~\cite{}. Therefore some type
annotations are necessary to guide type inference. In response to this 
challenge several decidable type systems requiring some annotations 
have been proposed~\cite{}. Two closely related type systems that 
support \emph{predicative} higher-ranked type inference were proposed 
by Peyton Jones et al.~\cite{} and Dunfield and
Krishnaswami (henceforth denoted as DK). 
These type systems are
popular among language designer and their ideas have been adopted by
several modern functional languages, including Haskell and PureScript.
In those type systems
type annotations are required for polymorphic arguments of functions.
but other type annotations can be omitted. A canonical example is:
\bruno{the following text needs to be changed, since it is basically
  as in ITP.}
\begin{verbatim}
hpoly = (\f :: forall a. a -> a) -> (f 1, f 'c')
\end{verbatim}
The function \verb|hpoly| cannot be
type-checked in Hindley-Milner.  The type of \verb|hpoly| is 
\verb|(forall a. a -> a) -> (Int, Char)|. The single universal quantifier
does not appear at the top-level. Instead it is used to quantify a
type variable \verb|a| used in the first argument of the
function.
Notably \verb|hpoly| requires a type annotation for the first
argument (\verb|forall a. a -> a|). 
Despite these additional annotations,
the type-inference algorithm employed by GHC Haskell~\cite{jones2007practical} preserves 
many of the desirable properties of Hindley-Milner.

Despite the importance of type inference in practice and also in
academic research, there is little work in mechanically formalizing
such advanced forms of type inference in theorem provers. This is at
odds with a current trend in programming language research. In
particular both the POPLMark challenge~\cite{} and CompCert ~\cite{}
have significantly promoted the use of theorem provers to model
variaous aspects of programming languages. Indeed today there are
routinely papers in top conferences that use theorem provers to
mechanically formalize various aspects of programming languages,
including: \emph{dynamic and static semantics} and their
correctness properties~\cite{}, \emph{compiler correctness}~\cite{},
\emph{correctness of optimizations}~\cite{} or even certain forms of
\emph{abstract interpretation}. Mechanical formalizations have
numerous advantages~\cite{read to make sure I do not miss any!}.

>>>>

As functional languages evolved the need for more expressive
power has motivated language designers to look beyond Hindley-Milner.
In particular one popular direction has been to allow 
\emph{higher-ranked polymorphism} where polymorphic types can
occur anywhere in a type signature.  
An important challenge is that full type inference for higher-ranked
polymorphism is know to be undecidable~\cite{}. Therefore some type
annotations are necessary to guide type inference. In response to this 
challenge several decidable type systems requiring some annotations 
have been proposed~\cite{}. Two closely related type systems that 
support \emph{predicative} higher-ranked type inference were proposed 
by Peyton Jones et al.~\cite{} and Dunfield and
Krishnaswami (henceforth denoted as DK). 
These type systems are
popular among language designer and their ideas have been adopted by
several modern functional languages, including Haskell and PureScript.
In those type systems
type annotations are required for polymorphic arguments of functions,
but other type annotations can be omitted. A canonical example is:
\bruno{the following text needs to be changed, since it is basically
  as in ITP.}
\begin{verbatim}
hpoly = (\f :: forall a. a -> a) -> (f 1, f 'c')
\end{verbatim}
The function \verb|hpoly| cannot be
type-checked in Hindley-Milner.  The type of \verb|hpoly| is the rank-2 type:
\verb|(forall a. a -> a) -> (Int, Char)|. Notably (and unlike
Hindley-Milner) the lambda argument \verb|f| requires a
\emph{polymorphic} type annotation.
This annotation is needed because the single universal quantifier
does not appear at the top-level. Instead it is used to quantify a
type variable \verb|a| used in the first argument of the
function. 
Despite these additional annotations, Peyton Jones et al. and DK's
type inference algorithms preserve many of the desirable properties 
of Hindley-Milner. For example the applications of \verb|f| implicitly 
instantiate the polymorphic type arguments of \verb|f|.

Despite the importance of type inference in practice and also in
academic research, there is little work in mechanically formalizing
such advanced forms of type inference in theorem provers. This is at
odds with the current trend of mechanical formalizations in
programming language research. In particular both the POPLMark
challenge~\cite{} and CompCert ~\cite{} have significantly promoted
the use of theorem provers to model various aspects of programming
languages. Indeed today there are routinely papers in top conferences
that use theorem provers to mechanically formalize: \emph{dynamic and
  static semantics} and their correctness properties~\cite{},
\emph{compiler correctness}~\cite{}, \emph{correctness of
  optimizations}~\cite{}, certain forms of \emph{abstract
  interpretation} or proofs involving \emph{logical relations}. The
main argument for mechanical formalizations is a simple one. Proofs
for programming languages tend to be \emph{long}, \emph{tedious} and
\emph{error-prone}. In such proofs it is very easy to make mistakes
that invalidate the whole development. Furthermore, readers and
reviewers often do not have time to look at the proofs carefuly to
check their correctnessness. Therefore errors can go unnoticed for a
long time.  Mechanical formalizations provide, in principle, a natural
solution for such problems. Theorem provers can automaticaly check and
validate the proofs, which removes the burden from checking from both
the person doing the proofs as well as readers or reviewers.

DK problem in lemma ??. Castagna problem in a recent type inference paper.

with the exception of some work
done on the formalization of 
certain aspects of Hindley-Milner type inference, we do not know
